Assessing Ranking Fidelity of Competition Structures via Weighted Mutual Information

### 2025-10-17 CMSAC_presentation/Zach_slides.qmd

## Introduction

```{=html}
<!-- Overall comments 
I think the presentation is too long for 12 minutes (+ questions), but
I feel like I didn't cover everything we want. What is plausable to cover
in a 12-minute period? 
Also, we need to figure out the notation for the final metric because I think
the weight should have the subscript m, not I. And I probably was not consistent
with the usage of theta or calling it true rank/true strength. -->
```

-   There exists numerous competition formats across all professional sports
-   Which competition structure is best?
-   The "best" competition structure depends on how well it reflects the competitors' true strengths
-   We will propose a metric that quantifies how effectively different tournament structures convey information about competitors' true strengths

<!-- Should we talk about Efficacy vs Effectivity here -->

## Extreme Tournament

- 4 team single elimination tournament:
  - Dream Team vs Work Group vs Middle School Girls Team vs Babies
- If we only care about the top team winning, seeding structure does not matter
- If we care about who gets second and third, seeding becomes important
  - Would not pair up Work Group and Dream Team in the first round

## Comparing Structures

- Comparing different structures like round robin and single eliminations, which is best?

- Efficacy: Final rankings should match the true rankings of the strength of the competitors

- Effectivity: Interested in correctly ranking the best competitor as first

CITEE

## Efficacy and Effectivity

- Other competitions where the correct ranking of some, but not all, of the competitors is of import
  - Olympic competition ranks the top 3 competitors

- "Salient ranks": the ranks of a competition structure that matter
  - Efficacy: 1 through $n$ matter
  - Effectivity: 1 is only salient rank

- We will propose a metric that quantifies how effectively different tournament structures convey information about competitors’ true strengths

## Game Simulation

-   Given the assigned "known strength", $\theta$, we simulate a game using the Bradley-Terry Model

$$\Pr(i > j) = \frac{e^{\theta_i}}{e^{\theta_i} + e^{\theta_j}} = \frac{1}{1+e^{\theta_i - \theta_j}}$$

-   Each competition structure is simulated given the number of teams, $n$, and $\theta$
-   Below is an example of the output from one simulation of a single elimination tournament structure

```{r}
renamed_df <- all_results_norm %>% 
  select(-c(simulation, distribution, seed, teams)) %>%
  rename(theta = true_strength, r = true_rank, r_hat = rank_hat) %>%
  mutate(theta = round(theta,3))

head(renamed_df,8)
```

@@

-   If one views the true ranking of the teams in a tournament as a message to be sent to a receiver and the outcome of the tournament as a message that is received, we can measure the “goodness” of a tournament in terms of its ability to accurately transmit the true ranking.

@@

<!-- -->

@@

-   A limitation of mutual information is that it does not distinguish between the direction of information
-   In other words, a ranking of $(1,2,3,4)$ is equivalent to $(4,3,2,1)$
-   Because of this, we proposed a weighting function $w(r, \hat{r})$
-   So, the new formula is: $$ wI_m(r,\hat{r}) = \sum_r \sum_{\hat{r}} w(r,\hat{r}) * p(r, \hat{r}) \log{\frac{p(r,\hat{r})}{p(r)p(\hat{r})}} $$

@@

-   In our original paper submission, we used squared error loss as the weighting function: $$ w(r, \hat{r}) =
    \begin{cases}
    \frac{1}{(r-\hat{r})^2}, & r \neq \hat{r} \\
    1, & r = \hat{r}
    \end{cases} $$

-   However, after comments and reviews from people at CSAS and CMSAC, we have decided to use a weighted Kendall's Tau

@@

$$ \tau_w = \frac{\sum_{i\neq j} W_{ij} \times \text{sgn}(i-j)\times \text{sgn}(R_i - R_j)}{\sum_{i,j} W_{ij} - \sum_i W_{ii}}$$

-   $W$ is a $n\times n$ lower-triangular matrix, where

<!-- Should this be capital W??? -->

$$W_{ij} = w_{\min(i,j)}, \text{ for } i,j = m,2m, \dots, n\times m$$ - $m$: vector of length $n$ representing the "salient weights", or the importance of correctly getting each ranking

-   If $\tau_w = 1$, then the ranking is identical to the true ranking; if $\tau_w = 0$, then the order of teams is perfectly inversely proportional to the true rankings

## Weighted Kendall Example

-   If $\hat{r} = (2,1,3,4)$ and m = $(3,1,1,0)$, then: $$ W = \begin{pmatrix}
    3 & 3 & 3 & 3 \\
    3 & 1 & 1 & 1 \\
    3 & 1 & 1 & 1 \\
    3 & 1 & 1 & 0
    \end{pmatrix} $$

$$\sum_{i,j} W_{ij} = 29; \sum_{i} W_{ii} = 5; \sum_{i\neq j} W_{ij} \times \text{sgn}(i-j)\times \text{sgn}(R_i - R_j) = 18 $$

-   So, $\tau_w$ for this example would be $\frac{18}{29-5} = 0.75$

@@

-   Mutual Information can be normalized by dividing by the joint entropy

-   Entropy: $-p(x)\log{p(x)}$

@@

-   $H(r)$ and $H(\hat{r})$ are theoretically equivalent, leaving $H(r,\hat{r}) = -2\sum_{r} p(r)\log(p(r)) = -2(n!) \times (1/n!) \times\log_2(1/n!) = \log_2((n!)^2)$ <!-- Is this true?? I can't remember if they are equivalent or their maximums were equivalent -->

-   So, the final formula for the metric is: $$ \frac{wI_m(r,\hat{r})}{H(r) + H(\hat{r})} = \frac{\sum_r \sum_{\hat{r}} w(r,\hat{r}) * p(r, \hat{r}) \log{\frac{p(r,\hat{r})}{p(r)p(\hat{r})}}}{\log_2((n!)^2)}$$

@@

<!-- Should these plots be on separate slides? -->

```{r out.width="49%", fig.show='hold', fig.align='center'}
knitr::include_graphics("tournament_comparison_plot.png")
knitr::include_graphics("tournament_comparison_plot_close.png")
```

>>>>

@@

::: notes
If one views the true ranking of the teams in a tournament as a message to be sent to a receiver and the outcome of the tournament as a message that is received, we can measure the "goodness" of a tournament in terms of its ability to accurately transmit the true ranking.
:::

@@

@@

-   A limitation of mutual information is that it does not distinguish between the direction of information
-   In other words, given equal probabilities, a ranking of $(1,2,3,4)$ is equivalent to $(4,3,2,1)$
-   Because of this, we proposed a new weighting function $w(r, \hat{r})$\
    $$ wI_m(r,\hat{r}) = \sum_r \sum_{\hat{r}} w(r,\hat{r}) * p(r, \hat{r}) \log{\frac{p(r,\hat{r})}{p(r)p(\hat{r})}} $$

@@

-   In our original paper submission, we used squared error loss as the weighting function: $w(r, \hat{r}) = \begin{cases} \frac{1}{(r-\hat{r})^2}, & r \neq \hat{r} \\ 1, & r = \hat{r} \end{cases}$
-   However, after comments and reviews from people at CSAS and CMSAC, we have decided to use a weighted Kendall's Tau

@@

$$ \tau_w = \frac{\sum_{i\neq j} W_{ij} \text{ sgn}(i-j) \text{ sgn}(R_i - R_j)}{\sum_{i,j} W_{ij} - \sum_i W_{ii}}$$

-   If $\hat{r} = (2,1,3,4)$, then:

$$\sum_{i\neq j} W_{ij} \text{ sgn}(i-j) \text{ sgn}(R_i - R_j) = 10$$

-   So, $\tau_w$ for this example would be $\frac{10}{15-3} = 0.833$

## Top K-teams Weighted Kendall

```{r}
rhat_df %>% 
  select(-squared_error) %>%   
  filter(permutation %in% c("1234", "4321", "1324", "2143")) %>%   
  arrange(permutation,k) %>%   
  rename("Weighted Kendall" = weighted_kendall)
```

$$ \tau_w = \frac{\sum_{i\neq j} W_{ij} \text{ sgn}(i-j) \text{ sgn}(R_i - R_j)}{\sum_{i,j} W_{ij} - \sum_i W_{ii}}$$

## Comparing Weighted Kendall's Tau to Squared Error

```{r}
rhat_df %>%    
  filter (k==4) %>%   
  select(-k) %>%   
  arrange(desc(weighted_kendall)) %>%    
  rename("Weighted Kendall"= weighted_kendall, "Squared Error" = squared_error)
```

@@

-   To make comparisons across different values of $n$, we normalize by the joint entropy

-   So, the final formula for the metric is:

$$\frac{wI_m(r,\hat{r})}{H(r,\hat{r})} = \frac{\sum_r \sum_{\hat{r}} w(r,\hat{r}) * p(r, \hat{r}) \log{\frac{p(r,\hat{r})}{p(r)p(\hat{r})}}}{\log_2((n!)^2)}$$

@@

-   Given the assigned "known strength", $\theta$, we simulate a game using the Bradley-Terry Model: $\Pr(i > j) = \frac{e^{\theta_i}}{e^{\theta_i} + e^{\theta_j}} = \frac{1}{1+e^{\theta_i - \theta_j}}$
-   Below is output from one simulation of a single elimination tournament structure

```{r}
renamed_df <- all_results_norm %>%    
  select(-c(simulation, distribution, seed, teams)) %>%   
  rename(theta = true_strength, r = true_rank, r_hat = rank_hat) %>%   
  mutate(theta = round(theta,3))  
head(renamed_df,8)
```

@@

```{r out.width="49%", fig.show='hold', fig.align='center'}
knitr::include_graphics("tournament_comparison_plot.png")
```

## Close-Up Results

```{r out.width="49%", fig.show='hold', fig.align='center'}
knitr::include_graphics("tournament_comparison_plot_close.png")
```
